---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# webterminal

<!-- badges: start -->
[![R-CMD-check](https://github.com/sondreskarsten/webterminal/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sondreskarsten/webterminal/actions/workflows/R-CMD-check.yaml)
[![integration-test](https://github.com/sondreskarsten/webterminal/actions/workflows/integration-test.yaml/badge.svg)](https://github.com/sondreskarsten/webterminal/actions/workflows/integration-test.yaml)
<!-- badges: end -->

A fast, browser-based terminal for RStudio Server. Launches a
[ttyd](https://github.com/tsl0922/ttyd) or
[shellinabox](https://github.com/shellinabox/shellinabox) daemon on
localhost and opens it in the RStudio Viewer pane through the built-in
`/p/<hex>/` reverse proxy — no new ports, no outbound connections, no
SSH.

## Why

RStudio Server's built-in terminal pane routes keystrokes through the
IDE's RPC subsystem, adding noticeable input lag. webterminal bypasses
this entirely: ttyd's native WebSocket transport connects xterm.js
directly to a pty, giving you a terminal that feels like a local SSH
session.

| Feature | RStudio terminal | webterminal (ttyd) |
|---|---|---|
| Transport | RPC / WebSocket fallback | Native WebSocket |
| Copy/paste | Broken in some setups | Native browser |
| Persistent sessions | No | Yes (tmux) |
| Survives R restart | No | Yes |
| New ports needed | No | No |

## Installation

```{r install-github, eval = FALSE}
# install.packages("pak")
pak::pak("sondreskarsten/webterminal")
```

### System dependencies

Install once on the RStudio Server host:

```bash
# Ubuntu / Debian
sudo apt-get install ttyd tmux iproute2

# Fedora / RHEL
sudo dnf install ttyd tmux iproute

# macOS (Homebrew)
brew install ttyd tmux
```

See `vignette("installing-ttyd")` for per-distro details, corporate
Nexus mirrors, and Docker recipes.

## Quick start

```{r quick-start, eval = FALSE}
library(webterminal)

# Check your setup
webterminal_doctor()

# Open a terminal in the RStudio Viewer
web_terminal()
```

The terminal opens in the Viewer pane. Click **Show in new window** to
pop it out into a full browser tab.

## Backends

webterminal ships three backends. Choose by name:

```{r backends}
library(webterminal)
terminal_backends()
```

| Backend | Transport | Persistent | Best for |
|---|---|---|---|
| `ttyd` | WebSocket | No | Quick throwaway shells |
| `ttyd-tmux` | WebSocket | Yes | Daily work — survives tab close |
| `shellinabox` | AJAX | Yes | Fallback when ttyd unavailable |

```{r choose-backend, eval = FALSE}
web_terminal("ttyd-tmux")
```

## Daemon lifecycle

Daemons are detached processes — they survive R session restarts,
`rstudio-server` restarts, and browser tab closures.

```{r lifecycle, eval = FALSE}
# Start explicitly
terminal_start("ttyd-tmux")

# Check what's running
terminal_status()

# Open in viewer (daemon already up — instant)
web_terminal("ttyd-tmux")

# Stop explicitly
terminal_stop("ttyd-tmux")
```

## Auto-start on login

webterminal never writes to your `.Rprofile`. It prints a snippet you
paste yourself:

```{r snippet}
webterminal_snippet("ttyd-tmux")
```

## How it works

```
Browser tab
    │
    ▼
ALB / reverse proxy (existing)
    │
    ▼
rserver (port 8787)
    │   validates auth cookie + port token
    │   forwards via /p/<hex>/ proxy
    ▼
ttyd (127.0.0.1:<port>)
    │   xterm.js ↔ pty via WebSocket
    ▼
bash / tmux session
```

1. `terminal_start()` launches ttyd bound to `127.0.0.1` via the
   bundled shell script in `inst/scripts/`.
2. `web_terminal()` calls `rstudioapi::viewer()` with the localhost
   URL.
3. RStudio Server's `rserver` process (controlled by
   `www-proxy-localhost=1` in `/etc/rstudio/rserver.conf`) rewrites
   this to a `/p/<hex>/` proxy URL and forwards HTTP + WebSocket
   Upgrade to the daemon.
4. The terminal renders in your browser — native copy/paste, full
   color, no IDE overhead.

See `vignette("architecture")` for the full technical deep-dive
including security model, cross-user isolation, and load balancer
configuration.

## Configuration

| Option / env var | Purpose | Default |
|---|---|---|
| `options(webterminal.backend = "ttyd-tmux")` | Default backend | `"ttyd"` |
| `options(webterminal.port = 7681L)` | Override base port | Deterministic from `$USER` |
| `WEBTERMINAL_PORT` | Override base port (env var) | — |
| `WEBTERMINAL_TTYD` | Force ttyd binary path | `Sys.which("ttyd")` |
| `WEBTERMINAL_SHELLINABOXD` | Force shellinaboxd binary path | `Sys.which("shellinaboxd")` |

## Diagnostics

```{r doctor, eval = FALSE}
webterminal_doctor()
```

```
── webterminal doctor ───────────────────────
[ok] ttyd:         /usr/bin/ttyd
    version:      ttyd version 1.7.4
[!!] shellinaboxd: [NOT FOUND]
[ok] tmux:         /usr/bin/tmux
    version:      tmux 3.4
[ok] bash:         /usr/bin/bash
[ok] ss:           /usr/bin/ss
[ok] RStudio:      available

No daemons running.
─────────────────────────────────────────────
```

## Related work

- [servr](https://github.com/yihui/servr) — in-process HTTP server
  for static files (similar port-management patterns)
- [httpgd](https://github.com/nx10/httpgd) — HTTP graphics device
  (similar `rstudioapi::viewer()` integration)
- [rstudioapi](https://rstudio.github.io/rstudioapi/) — the API that
  makes the `/p/<hex>/` viewer proxy work

## License

MIT
