---
title: "Installing system dependencies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing system dependencies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

webterminal requires three system binaries that are **not installed
by the R package**:

| Binary | Purpose | Required? |
|---|---|---|
| `ttyd` | WebSocket-based terminal daemon | Yes (primary backend) |
| `tmux` | Terminal multiplexer for session persistence | Yes for `ttyd-tmux` and `shellinabox` backends |
| `ss` | Socket statistics for port probing | Recommended (falls back to `socketConnection()`) |

Optional:

| Binary | Purpose |
|---|---|
| `shellinaboxd` | AJAX-based terminal daemon (fallback when ttyd unavailable) |

## Ubuntu / Debian

### ttyd

ttyd is in the `universe` repository on Ubuntu 22.04+ and in Debian
`testing`/`unstable`:

```{bash}
sudo apt-get update
sudo apt-get install ttyd
```

Verify:

```{bash}
ttyd --version
# ttyd version 1.7.4
```

If your corporate mirror does not include the `universe` component,
ask your platform team to add it, or install from the upstream
release:

```{bash}
# Only if ttyd is not in your apt repos
curl -sL https://github.com/tsl0922/ttyd/releases/latest/download/ttyd.x86_64 \
  -o /usr/local/bin/ttyd
chmod +x /usr/local/bin/ttyd
```

### tmux and iproute2

Both are in `main` and universally available:

```{bash}
sudo apt-get install tmux iproute2
```

### shellinabox (optional)

shellinabox re-entered Debian `unstable` on 2025-10-17 (version
2.21-1). On Ubuntu Jammy (22.04):

```{bash}
sudo apt-get install shellinabox
```

Note: shellinabox uses AJAX long-polling rather than WebSocket. It is
noticeably slower than ttyd. Use it only when ttyd is unavailable.

### All at once

```{bash}
sudo apt-get install ttyd tmux iproute2
```

## RHEL / Fedora / CentOS

ttyd is not in the default RHEL/CentOS repos. Install from EPEL or
the upstream binary:

```{bash}
# Fedora
sudo dnf install ttyd tmux iproute

# RHEL/CentOS (upstream binary)
curl -sL https://github.com/tsl0922/ttyd/releases/latest/download/ttyd.x86_64 \
  -o /usr/local/bin/ttyd
chmod +x /usr/local/bin/ttyd
sudo dnf install tmux iproute
```

## macOS (Homebrew)

ttyd is in Homebrew core:

```{bash}
brew install ttyd tmux
```

Note: webterminal is designed for RStudio Server (Linux). On macOS it
works with RStudio Desktop but `rstudioapi::viewer()` opens a local
Viewer pane rather than proxying through `/p/<hex>/`.

## Alpine Linux

```{bash}
apk add ttyd tmux iproute2
```

## Docker containers

If your RStudio Server runs inside a Docker container, add to the
Dockerfile:

```{bash}
RUN apt-get update && apt-get install -y --no-install-recommends \
    ttyd tmux iproute2 \
  && rm -rf /var/lib/apt/lists/*
```

## Corporate environments with Nexus mirrors

If your apt mirror is a Nexus proxy that only shadows specific
components, check whether `universe` is included:

```{bash}
apt-cache policy ttyd
```

If it returns `Unable to locate package ttyd`, the mirror is missing
the `universe` component. Either:

1. Ask your platform team to add `universe` to the Nexus proxy, or
2. Download the static binary directly (see Ubuntu section above), or
3. Use `shellinabox` from `universe` (it entered `universe` earlier
   than `ttyd` on some mirrors).

## Verifying the installation from R

After installing the system binaries, verify from R:

```{r}
library(webterminal)
webterminal_doctor()
```

Expected output:

```
-- webterminal doctor -----------------------
[ok] ttyd:         /usr/bin/ttyd
    version:      ttyd version 1.7.4
[!!] shellinaboxd: [NOT FOUND]
[ok] tmux:         /usr/bin/tmux
    version:      tmux 3.4
[ok] bash:         /usr/bin/bash
[ok] ss:           /usr/bin/ss
[ok] RStudio:      available

No daemons running.
--------------------------------------------
```

`shellinaboxd [NOT FOUND]` is fine — it is optional. All other lines
should show `[ok]`.

## Disabling the system shellinabox service

On Ubuntu, `apt-get install shellinabox` enables a system service
bound to `0.0.0.0:4200`. webterminal manages its own shellinabox
instance bound to `127.0.0.1`, so you should disable the system
service:

```{bash}
sudo systemctl disable --now shellinabox
sudo systemctl mask shellinabox
```

This is a one-time step. webterminal's daemon launches via
`terminal_start("shellinabox")` on a user-specific port, bound to
loopback only.

## Disabling the system ttyd service

Similarly, some distributions enable a system ttyd service on install.
Disable it:

```{bash}
sudo systemctl disable --now ttyd
sudo systemctl mask ttyd
```

webterminal launches its own ttyd instances on user-specific ports.

## Overriding binary paths

If ttyd is installed to a non-standard location, set the environment
variable before loading the package:

```{r}
Sys.setenv(WEBTERMINAL_TTYD = "/opt/custom/bin/ttyd")
library(webterminal)
webterminal_doctor()  # should show "NOTE: ttyd path was forced by WEBTERMINAL_TTYD"
```

Or in `.Renviron`:

```
WEBTERMINAL_TTYD=/opt/custom/bin/ttyd
```
