---
title: "Getting started with webterminal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with webterminal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What webterminal does

webterminal opens a fast, browser-based terminal inside the RStudio
Server viewer pane. It works by launching a lightweight web-terminal
daemon (`ttyd` or `shellinabox`) on localhost and letting RStudio
Server's built-in reverse proxy (`/p/<hex>/`) forward the connection
to your browser.

The result is a terminal tab that feels native — WebSocket transport,
real copy/paste, and optional `tmux` persistence across browser
reconnects.

## Prerequisites

webterminal is a connector, not an installer. You need three system
binaries present on the RStudio Server host before the package can do
anything useful:

```{bash}
sudo apt-get install ttyd tmux iproute2
```

See `vignette("installing-ttyd")` for detailed per-distro instructions.

## Quick start

```{r}
library(webterminal)

# Check what's installed and what's missing
webterminal_doctor()
```

The doctor report shows each binary's path and version, whether
RStudio Server is detected, and any running daemons. If something is
missing, it prints the exact install command.

### Open a terminal

```{r}
web_terminal()
```
This does three things in sequence:

1. Checks whether a `ttyd` daemon is already listening on your
   deterministic port.
2. If not (and you're in an interactive session), launches one via the
   bundled `inst/scripts/start-ttyd.sh`.
3. Calls `rstudioapi::viewer()` with the localhost URL, which RStudio
   Server proxies to your browser.

The terminal opens in the Viewer pane. Click the "Show in new window"
icon to pop it out into a full browser tab.

### Choose a backend

webterminal ships three backends:

```{r}
terminal_backends()
```

| Backend | Transport | Persistent | Use case |
|---|---|---|---|
| `ttyd` | WebSocket | No | Lightweight, ephemeral shell |
| `ttyd-tmux` | WebSocket | Yes | Survives tab close / reconnect |
| `shellinabox` | AJAX | Yes | Fallback when ttyd unavailable |

Switch backends by name:

```{r}
web_terminal("ttyd-tmux")
```

Or set a default for the session:

```{r}
options(webterminal.backend = "ttyd-tmux")
web_terminal()
```

## Explicit lifecycle management

For finer control, use `terminal_start()` and `terminal_stop()`
directly:

```{r}
# Start the daemon (survives R session restart)
terminal_start("ttyd-tmux")

# Check what's running
terminal_status()

# Open the viewer (daemon already up, so instant)
web_terminal("ttyd-tmux")

# Later: stop the daemon
terminal_stop("ttyd-tmux")
```

Daemons started by webterminal are **detached processes** — they
survive R session restarts, RStudio Server restarts, and browser tab
closures (when using `ttyd-tmux`). They must be stopped explicitly
with `terminal_stop()`.

The daemon registry is persisted to
`tools::R_user_dir("webterminal", "cache")`, so `terminal_status()`
can find daemons started in a previous R session.

## Multiple backends simultaneously

Each backend binds to a different port (deterministic, based on your
username). You can run all three at once:
```{r}
terminal_start("ttyd")
terminal_start("ttyd-tmux")
terminal_status()
```

## Auto-start on RStudio Server login

webterminal never modifies your `.Rprofile`. Instead, it prints a
snippet you can paste yourself:

```{r}
webterminal_snippet("ttyd-tmux")
```

This prints:

```r
if (interactive() && requireNamespace("rstudioapi", quietly = TRUE) &&
    rstudioapi::isAvailable()) {
  try(webterminal::terminal_start("ttyd-tmux"), silent = TRUE)
}
```

Paste it into `~/.Rprofile` (open with `usethis::edit_r_profile()`)
and every future RStudio session will have a terminal daemon ready.

## Port allocation

webterminal computes a deterministic base port from your username:

```
base = 4200 + (sum of UTF-8 codepoints in $USER) mod 90
```

Each backend adds a fixed offset (0, 1, 2), so different backends
never collide. Override the base port with:

```{r}
# Per-session
options(webterminal.port = 7681L)

# Per-environment (e.g. in .Renviron)
Sys.setenv(WEBTERMINAL_PORT = "7681")
```

## Viewer modes

```{r}
# Default: RStudio viewer if available, else system browser
web_terminal()

# Force RStudio viewer (errors outside RStudio)
web_terminal(viewer = "rstudio")

# Force system browser
web_terminal(viewer = "browser")

# Just return the URL (for scripting / piping)
url <- web_terminal(viewer = "url")
```

## Environment variables

| Variable | Purpose |
|---|---|
| `WEBTERMINAL_PORT` | Override base port (takes precedence over option) |
| `WEBTERMINAL_TTYD` | Force path to `ttyd` binary |
| `WEBTERMINAL_SHELLINABOXD` | Force path to `shellinaboxd` binary |

When a binary path is forced, `webterminal_doctor()` reports it:

```
NOTE: ttyd path was forced by WEBTERMINAL_TTYD
```

## Troubleshooting

Run `webterminal_doctor()` first. Common issues:

**"ttyd not found on PATH"**
: Install ttyd. See `vignette("installing-ttyd")`.

**"Daemon launched but port not responding after 3 seconds"**
: The bundled launch script failed silently. Check the log at
  `/tmp/webterminal-ttyd-<port>.log`.

**Viewer pane shows a blank page**
: The RStudio Server localhost proxy may not support WebSocket
  upgrade. Check `www-proxy-localhost=1` in
  `/etc/rstudio/rserver.conf`. See `vignette("architecture")`.

**Terminal opens but immediately disconnects**
: Your load balancer's idle timeout is too short. Set `--ping-interval`
  in the launch script below the ALB timeout (default is 20 seconds).
